Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-12-13 Chapter_1/04_Manuscript/Manuscript.tex
Various updates

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage[colorinlistoftodos]{todonotes}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

%------------------------------------------------------------------------------
% Short Version
%------------------------------------------------------------------------------
% The ability to disperse is contingent a sufficient degree of landscape
% connectivity, which is why the identification and preservation of movement
% corridors that promote connectivity has become a task of extraordinary
% importance. Currently, ecologists mainly rely on least-cost analysis or circuit
% theory to investigate connectivity, albeit both methods make several assumptions
% are hardly met in reality. To address these issues, simulations from
% individual-based movement models have been proposed, yet a unified framework to
% simulate dispersal and quantify connectivity is lacking.
%
% Here, we employ a simple three-step workflow that combines several
% already-existing methods to assess connectivity using simulated dispersal
% trajectories. In the first step, we use integrated step selection functions to
% parametrize a mechanistic movement model rendering dispersal behavior. In a
% second step, we apply the parametrized model as an individual-based movement
% model to simulate dispersal trajectories. Finally, in the third step, we
% consolidate simulated trajectories into three complementary connectivity maps,
% each focusing on a different aspect of landscape connectivity.
%
% We apply these steps to assess landscape connectivity for dispersing African
% wild dogs (\textit{Lycaon pictus}) southern Africa. We thereby demonstrate that
% simulations from integrated step selection functions offer a simple, yet
% powerful alternative to traditional connectivity modeling techniques.

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of dispersal & connectivity and mow its measured
Dispersal of individuals is crucial for long-term species persistence and
depends on a sufficient degree of landscape connectivity. Thusfar, researchers
have investigated connectivity primarily using least-cost analysis and circuit
theory, although both methods make assumptions that are rarely met in reality.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the least costly route to reach it. Circuit theory relies on
the assumption of a complete random walk without any directional persistence
along the dispersing route. Albeit these shortcomings can be overcome by
spatio-temporally explicitly simulating dispersing trajectories across the
landscape, a unified approach for such a simulation is lacking.

% What we do
Here, we present a simple three-step workflow to simulate dispersal and assess
connectivity starting from empirical GPS movement data. To exemplify its
application, we employ the workflow to dispersal data collected on the
endangered African wild dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi
Transfrontier Conservation Area, the world’s largest transboundary conservation
area. In step one, we use integrated step selection functions to fit a
mechanistic movement model describing habitat and movement preferences of
dispersers. In step two, we apply the parameterized model to simulate individual
dispersal trajectories. In step three, we compress the simulated trajectories
into three complementary connectivity maps: a heatmap that highlights frequently
traversed areas, a betweenness map that pinpoints dispersal corridors, and a map
of interpatch connectivity, indicating the presence or absence of functional
links between habitat patches.

% Some results
Our results show that dispersing African wild dogs move with directional
persistence in the vicinity to waterbodies and preferrably through areas with
little human influence. This is reflected in the simulation with resulted in
several dispersal hotspots and corridors along waterbodies through areas with
little or no anthropogenic pressure. Connectivity between national parks inside
the KAZA-TFCA appears to be good, albeit the parks in Zambia and Zimbabwe are
comparably less well connected to the rest of KAZA-TFCA, suggesting the
necessity for additional protected areas that serve as ``stepping stones'' into
these regions.

% Case study
We show that simulations from step-selection functions offer a simple yet
powerful alternative to traditional connectivity modeling techniques, although
necessitating informed decisions about the number of simulated individuals and
the duration of simulated dispersal events. Ultimately, our workflow permits a
more mechanistic understanding of dispersal behaviour and landscape connectivity
making it useful for a variety of applications in ecological, evolutionary, and
conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the ability
to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. The identification of dispersal
corridors not only necessitates a comprehensive understanding of the factors
that limit dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two most
commonly used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based methods that quantify conductance of the
landscape based on habitat permeability \citep{Zeller.2012}. However, both
approaches rest on assumptions that appear unsuitable for dispersers, which is
why  simulating dispersal from individual-based movement models provides a
promising alternative approach to modeling connectivity \citep{Diniz.2019}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be fulfilled by
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory and are therefore less aware of
associated movement costs and potential endpoints \citep{Koen.2014, Cozzi.2020}.
CT, on the contrary, posits that animals move according to a random walk,
entailing that autocorrelation between subsequent movements cannot be rendered
\citep{Diniz.2019}. For dispersers, however, autocorrelated movements are
regularly observed \citep{Cozzi.2020, Hofmann.2021}, implying that subsequent
relocations tend to fall onto a straight line. Moreover, graph-based methods are
unable to reflect the temporal dimension of dispersal, meaning that statements
about the approximate duration required to move between habitats impossible
\citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
trajectories from individual-based movement models (IBMMs) and by converting
simulated trajectories into meaningful measures of connectivity
\citep{Diniz.2019}). IBMMs allow to explicitly simulate how individuals move
across and interact with the encountered landscape \citep{Kanagaraj.2013,
Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020} and to render potential
interactions between movement behavior and habitat conditions. This shifts the
focus from a structural to a more functional view on connectivity
\citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}. Furthermore,
simulations from IBMMs generate movement sequentially, making the temporal
dimension of dispersal explicit and allowing to model autocorrelation between
subsequent movements \citep{Diniz.2019}. Finally, simulations from IBMMs do not
enforce connections towards preconceived endpoints, thereby mitigating biases
arising from misplaced endpoints. Despite these advantages, a unified framework
to simulate dispersal and assess connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
Here, we draw on recent advancements in the fields of movement ecology and
network theory to propose a simple three-step workflow for simulating dispersal
movements and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, we combine GPS movement data of dispersing individuals with relevant
habitat covariates to fit a mechanistic movement model using integrated step
selection functions (ISSFs, \citealp{Avgar.2016}). ISSFs permit
simultaneous inference on the focal species' habitat kernel (i.e. habitat
preferences), its movement kernel (i.e. movement preferences/capabilities), and
potential interactions between the two \citep{Avgar.2016, Fieberg.2021}. In step
two,  we simulate individual dispersal trajectories using the movement model
parametrized in step one. Finally, in step three, we convert simulated
trajectories into three complementary connectivity maps, each highlighting a
different aspect of connectivity. The proposed set of maps includes (i) a
heatmap revealing areas that are frequently traversed by simulated dispersers
(e.g. \citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map
delineating dispersal corridors and bottlenecks (e.g.
\citealp{BastilleRousseau.2018}), (iii) and a map of inter-patch connectivity,
depicting the presence or absence and intensity of use of specific connections,
and the average dispersal duration between corresponding habitat patches (e.g.
\citealp{Kanagaraj.2013}).

% Introduction of the study species
\subsection{Introduction of the Study Species}
The endangered African wild dog (\textit{Lycaon pictus}) represents a highly
mobile species whose persistence hinges on a  sufficient degree of landscape
connectivity. Once common throughout sub-Saharan Africa, this species has
disappeared from much of its historic range, largely due to human persecution,
habitat fragmentation, and disease outbreaks \citep{Woodroffe.2012}. Wild dogs
typically emigrate in single-sex coalitions \citep{McNutt.1996, Behr.2020}
capable of dispersing several hundred kilometers \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020}. Although previous studies have investigated
connectivity using LCPA \citep{Hofmann.2021} or CT \citep{Brennan.2020}, a more
comprehensive and mechanistic understanding of connectivity is missing for this
species (but see \citealp{Creel.2020}). With fewer than 6,000 free-ranging wild
dogs remaining in fragmented populations \citep{Woodroffe.2012}, reliable
information on landscape connectivity is essential for the conservation of this
endangered carnivore.

% % Some Background on SSFs
% \subsection{Some Background on SSFs}
% To fit a movement model, we propose to employ the framework of step selection
% functions (SSFs; \citealp{Fortin.2005, Thurfjell.2014, Avgar.2016}). Originally,
% SSFs were developed to learn about habitat preferences of the focal species
% using GPS data \citep{Fortin.2005}. The method works by comparing covariates
% along realized steps (the straight-line connecting two consecutive GPS
% relocations; \citealp{Turchin.1998}) to the same covariates along randomly
% generated alternative steps \citep{Fortin.2005, Thurfjell.2014} using
% conditional logistic regression \citep{Fortin.2005, Muff.2020}. Recently,
% \cite{Avgar.2016} have introduced \textit{integrated} SSFs (ISSFs), which is a
% generalization of regular SSFs and requires that movement metrics are included
% in the corresponding regression model. This not only reduces biases in estimated
% preferences, but also permits simultaneous inference on a species' habitat
% kernel (i.e. habitat preferences), its movement kernel (i.e. movement
% preferences/capabilities), and potential interactions between the two kernels
% \citep{Avgar.2016, Fieberg.2021}. Importantly, a model that was fitted using
% ISSFs resembles a fully mechanistic IBMM, based on which movement can be
% simulated \citep{Avgar.2016, Signer.2017}. Previously, ISSFs have been applied
% to simulate utilization distributions of resident animals \citep{Avgar.2016,
% Signer.2017}, yet the method remains underutilized for the simulation of
% dispersal and detection of movement corridors.

% % Some Background on Connectivity "Metrics"
% \subsection{Some Background on  Connectivity ``Metrics'' }
% Irrespective of the chosen modeling framework, simulated dispersal trajectories
% need to be processed into measures of connectivity. To date, the most common
% practice to translate simulations into connectivity measures is to compute
% inter-patch connectivity between distinct habitats \citep{Diniz.2019}.
% Inter-patch connectivity is a simple ratio between the number of dispersers
% successfully moving between habitat patches and the total number of simulated
% individuals from each patch (e.g. \citealp{Kanagaraj.2013}). As an alternative,
% some researchers have tallied overlapping trajectories into what might be called
% ``heatmaps'', i.e. maps that depict the frequency at which each spatial unit in
% the study area was traversed by simulated dispersers (e.g. \citealp{Peer.2008,
% Hauenstein.2019, Zeller.2020}). More recently, \cite{BastilleRousseau.2018}
% proposed a method to convert trajectories into a network, so that metrics from
% network theory can be computed. With regards to connectivity, the betweenness
% metric seems promising, as it indicates the number of shortest paths going
% through each node in the network and therefore highlights areas of exceptional
% importance for connecting different regions in the study area. Because each of
% the three metrics metrics focuses on a very different aspect of landscape
% connectivity, we propose to view them as complementary and to use them in
% concert when assessing connectivity.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract2.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis as proposed
    in this article. First, GPS data and habitat covariates have to be
    collected. The combined data is then analyzed in an integrated step
    selection model, which enables the parametrization of the focal species'
    habitat and movement kernels and results in a mechanistic movement model.
    The parametrized model is then treated as an individual-based movement model
    and used to simulate dispersal trajectories. Ultimately, simulated
    trajectories serve to produce a set of maps that are pertinent to landscape
    connectivity. This includes a heatmap, indicating the traversal frequency
    across each spatial unit of the study area, a betweenness map, highlighting
    movement corridors and bottlenecks, and, finally, an inter-patch
    connectivity map, where the frequency of connections and their average
    duration can be depicted.}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

% Case Study
\subsection{Case Study}
In this study, we illustrate the proposed three-step workflow
(\Cref{GraphicalAbstract}) to assess landscape connectivity for African wild
dogs within the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), the
world's largest transboundary conservation area. To this end, we use GPS
movement data of 16 dispersing wild dogs originating from a free-ranging
population in northern Botswana to parametrize a mechanstic movement model,
which we then employed to simulate 80,000 dispersal trajectories across the
landscape of the KAZA-TFCA. We anticipated that simulations based on our
three-step workflow would overcome several of the above highlighted conceptual
shortcomings of traditional connectivity models and provide a more comprehensive
view of landscape connectivity.

\section{Methods}
\subsection{Study Area}
The study area centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) is located in southern Africa (\Cref{StudyArea}a) and spans
more than 1.3 Mio. km\textsuperscript{2}, encompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA comprises parts of Angola, Botswana, Namibia,
Zimbabwe, and Zambia and hosts diverse landscapes; ranging from savannah to
grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydrogeographical feature and the world's largest
flood-pulsing inland delta. Although large portions of the KAZA-TFCA are part of
designated national parks and other protected areas, considerable human
influence from roads, agricultural sites, and settlements remains.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, blue] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, blue] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    comprised parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its main purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green).}
    \label{StudyArea}
  \end{center}
\end{figure}

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage[colorinlistoftodos]{todonotes}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
%------------------------------------------------------------------------------
% Short Version
%------------------------------------------------------------------------------
\textbf{Short Version }
The ability to disperse is contingent on a sufficient degree of landscape
connectivity, which is why the identification and preservation of movement
corridors that promote connectivity has become a task of extraordinary
importance. Currently, ecologists rely on least-cost analysis or circuit
theory to investigate connectivity, albeit both methods make several
assumptions that are hardly met in reality. To address these issues,
simulations from individual-based movement models have been proposed, yet a
unified framework to simulate dispersal and quantify connectivity is lacking.

Here, we propose a simple three-step workflow that combines several
already-existing methods to assess connectivity using simulated dispersal
trajectories. In step one, we use integrated step selection functions to
parametrize a mechanistic movement model rendering dispersal behavior. In step
two, we apply the parametrized model as an individual-based movement model to
simulate dispersal trajectories. In step three, we combine simulated
trajectories into three complementary connectivity maps, each focusing on a
different aspect of landscape connectivity.

We showcase the application of the proposed workflow using data of dispersing
African wild dogs (\textit{Lycaon pictus}) and assess landscape connectivity
within the world's largest transboundary conservation area in Southern Africa.
We thereby shed light into dispersing wild dogs' habitat and movement
preferences, while also uncovering crucial dispersal corridors. With this
analysis, we demonstrate that simulations from integrated step selection
functions offer a simple, yet powerful alternative to traditional connectivity
modeling techniques.
\end{abstract}

\newpage
\begin{abstract}
%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
\textbf{Long Version }
Dispersal of individuals is crucial for long-term species persistence and
depends on a sufficient degree of landscape connectivity. To date, researchers
primarily investigate connectivity using least-cost analysis and circuit theory,
both methods that make assumptions which are hardly met in reality. Least-cost
analysis assumes that animals move towards a known endpoint and are
knowledgeable about the most favorable route to reach it. Circuit theory
presumes a complete random walk and therefore fails to incorporate directional
persistence. These shortcomings can be overcome by spatio-temporally explicitly
simulating dispersal, yet a unified approach for such simulations is lacking.

Here, we present a simple three-step workflow to simulate dispersal and assess
connectivity starting from empirical GPS movement data. In step one, we use
integrated step selection functions to fit a mechanistic movement model
describing habitat and movement preferences of dispersers. In step two, we apply
the parameterized model to simulate individual dispersal trajectories. In step
three, we combine the simulated trajectories into three complementary
connectivity maps: a heatmap that highlights frequently traversed areas, a
betweenness map that pinpoints dispersal corridors, and a map of inter-patch
connectivity that indicates the presence and intensity of functional links
between habitat patches. As a case study, we analyse data on the endangered
African wild dog and assess landscape connectivity in the Kavango-Zambezi
Transfrontier Conservation Area, the world’s largest transboundary conservation
area.

Our results show that wild dogs preferably disperse with directional persistence
in the vicinity to water bodies and through areas with little human influence and
sparse woodland cover. Dispersal simulations and subsequent connectivity maps
revealed several dispersal hotspots and corridors across the extent of the
KAZA-TFCA. Connectivity between NPs inside the KAZA-TFCA was good, yet few
dispersers successfully moved from Zambia's NPs into neighboring areas.

We show that simulations from step-selection functions offer a simple yet
powerful alternative to traditional connectivity modeling techniques. In
contrast to traditional connect modeling techniques, simulations not only make
fewer unrealistic assumptions but also permit a more mechanistic understanding
of dispersal and landscape connectivity. This makes our workflow useful for a
variety of applications in ecological, evolutionary, and conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. Importantly, the
ability to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. The identification of dispersal
corridors not only necessitates a comprehensive understanding of the factors
that limit dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two most
commonly used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based methods that quantify conductance of the
landscape based on habitat permeability \citep{Zeller.2012}. However, both
approaches rest on assumptions that appear unsuitable for dispersers, which is
why simulating dispersal from individual-based movement models provides a
promising alternative approach to investigate connectivity \citep{Diniz.2019}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be fulfilled by
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory, so the associated movement costs and
potential endpoints are unknown to them \citep{Koen.2014, Cozzi.2020}. CT, on
the contrary, posits that animals move according to a random walk, entailing
that autocorrelation between subsequent movements cannot be rendered
\citep{Diniz.2019}. For dispersers, however, autocorrelated movements are
regularly observed \citep{Cozzi.2020, Hofmann.2021}, implying that dispersal
trajectories are usually strongly directional. Moreover, graph-based methods are
unable to reflect the temporal dimension of dispersal, so that statements about
the expected duration for moving between habitats are impossible
\citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
from individual-based movement models (IBMMs) and converting simulated
trajectories into meaningful measures of connectivity \citep{Diniz.2019}). In
contrast to graph-based methods, IBMMs allow to explicitly simulate how
individuals move across and interact with the encountered landscape
\citep{Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020}, as
well as to render potential interactions between movement behavior and habitat
conditions, thus shifting the focus from a structural to a more functional view
on connectivity \citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}.
Furthermore, because simulations from IBMMs generate movement sequentially, the
temporal dimension of dispersal becomes explicit and allows to model
autocorrelation between subsequent movements \citep{Diniz.2019}. Finally,
simulations from IBMMs do not enforce connections towards preconceived
endpoints, thereby mitigating biases arising from misplaced endpoints. Despite
these advantages, a unified framework to simulate dispersal and assess
connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
Here, we draw on recent advancements in movement ecology and propose a simple
three-step workflow for simulating dispersal and assessing landscape
connectivity (\Cref{GraphicalAbstract}). In step one, we combine GPS movement
data of dispersing individuals with relevant habitat covariates to fit a
mechanistic movement model using integrated step selection functions (ISSFs,
\citealp{Avgar.2016}). ISSFs permit inference on the focal species' habitat
kernel (i.e. habitat preferences), its movement kernel (i.e. movement
preferences/capabilities), and potential interactions between the two
\citep{Avgar.2016, Fieberg.2021}. In step two, we simulate individual dispersal
trajectories using the parametrized movement from step one. Comparable
simulations have already been employed to estimate steady-state utilization
distributions of residents \citep{Potts.2013, Signer.2017} and to model
landscape connectivity, yet disregarding interdependencies between habitat and
movement kernels \citep{Clark.2015, Zeller.2020}. Finally, in step three, we
convert simulated trajectories into three complementary connectivity maps, each
highlighting a different aspect of connectivity. The set of maps includes (i) a
heatmap revealing areas that are frequently traversed by simulated dispersers
(e.g. \citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map
delineating dispersal corridors and bottlenecks (e.g.
\citealp{BastilleRousseau.2018}), (iii) and a map of inter-patch connectivity,
depicting the presence and intensity of use of specific connections, as well as
the average dispersal duration required to connect corresponding habitat patches
(e.g. \citealp{Gustafson.1996, Kanagaraj.2013}).

% Introduction of the study species
\subsection{Case Study}
We showcase the application of the proposed workflow (\Cref{GraphicalAbstract})
using dispersal data collected on the endangered African wild dog
(\textit{Lycaon Pictus}). The African wild dog represents a highly mobile
species whose persistence hinges on a sufficient degree of landscape
connectivity. Once common throughout sub-Saharan Africa, this species has
disappeared from much of its historic range, largely due to human persecution,
habitat fragmentation, and disease outbreaks \citep{Woodroffe.2012}. Wild dogs
typically disperse in single-sex coalitions \citep{McNutt.1996, Behr.2020} and
are capable of dispersing several hundred kilometers \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020}. Although previous studies have investigated
connectivity using LCPA \citep{Hofmann.2021} or CT \citep{Brennan.2020}, a more
comprehensive and mechanistic understanding of connectivity is missing for this
species (but see \citealp{Creel.2020}). With fewer than 6,000 free-ranging wild
dogs remaining in fragmented populations \citep{Woodroffe.2012}, reliable
information on landscape connectivity is essential for the conservation of this
endangered carnivore. Here, we use GPS data of 16 dispersing wild dogs
originating from a free-ranging population in northern Botswana to parametrize a
mechanistic movement model, which we then employed to simulate 80,000 dispersal
trajectories across the landscape of the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). We anticipated that simulations based on our three-step workflow
would overcome several of the above highlighted conceptual shortcomings of
traditional connectivity models and provide a more comprehensive view of
landscape connectivity

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract2.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis as proposed
    in this article. First, GPS data and habitat covariates must be
    collected. The combined data is then analyzed in an integrated step
    selection model, which enables the parametrization of the focal species'
    habitat and movement kernels and results in a mechanistic movement model.
    The parametrized model is then treated as an individual-based movement model
    and used to simulate dispersal trajectories. Ultimately, simulated
    trajectories serve to produce a set of maps that are pertinent to landscape
    connectivity. This includes a heatmap, indicating the traversal frequency
    across each spatial unit of the study area, a betweenness map, highlighting
    movement corridors and bottlenecks, and, finally, an inter-patch
    connectivity map, where the frequency of connections and their average
    duration can be depicted.}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Study Area}
The study area centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) was located in southern Africa (\Cref{StudyArea}a) and
spanned more than 1.3 Mio. km\textsuperscript{2}, encompassing the entire
KAZA-TFCA (\Cref{StudyArea}b). The KAZA-TFCA comprises parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia and hosts a rich diversity of
landscapes, ranging from savannah to grassland and from dry to moist woodland
habitats. In its center lies the Okavango Delta, a dominant hydro-geographical
feature and the world's largest flood-pulsing inland delta. Large portions of
the KAZA-TFCA are part of designated national parks (NPs) and other protected
areas, yet considerable human influence through roads, agricultural sites, and
settlements remains.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, blue] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, blue] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA
    which comprises parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia.
    (b) The KAZA-TFCA currently represents the world's largest terrestrial
    transfrontier conservation area, covering a total area of 520'000
    km\textsuperscript{2}. Its main purpose is to re-establish connectivity
    between already-existing NPs (dark green) and other protected areas (light
    green).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 dispersing wild dogs (7 females and 9
males) between 2011 and 2019 from a free-ranging population in northern Botswana
(details on the data collection can be found in \cite{Cozzi.2020} and
\cite{Hofmann.2021}). Because behavior during dispersal is more pertinent to
landscape connectivity than behavior during residence \citep{Elliot.2014,
Abrahms.2017}, we discarded data collected when individuals were resident. We
determined the exact timing of emigration and settlement using direct field
observations and through visual inspection of the net squared displacement (NSD)
metric. NSD measures the squared Euclidean distance of a GPS relocation to a
reference point \citep{Borger.2012}, which we set to the center of each
dispersers' natal home range. Thus, dispersal was deemed to have started once an
individual left its natal home range and ended once the NSD metric remained
constant, indicating settlement. During dispersal, GPS collars recorded a fix
every 4 hours and regularly transmitted data over the Iridium satellite system.
To ensure regular timespans between GPS fixes, we removed any fix that was not
successfully obtained on the aspired 4-hourly schedule (allowing for a tolerance
of \( \pm \) 15 minutes). We then converted the remaining fixes (n = 4'169) into
steps, where each step represented the straight-line movement between two
consecutive GPS fixes \citep{Turchin.1998}.

\subsubsection{Habitat Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textit{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To correctly render seasonality in
water-cover, we applied a remote sensing algorithm that enabled us to obtain
weekly updated raster-layers for water-cover and corresponding layers for
distance-to-water from MODIS satellite imagery \citealp{Wolski.2017,
Hofmann.2021}. This algorithm is now implemented in the \textsf{floodmapr}
package which is available on GitHub
(\url{https://github.com/DavidDHofmann/floodmapr}). Using the frequently updated
floodmpas we were able to correctly assign the most appropriate covariate layer
to each observed movement step. To ensure a consistent resolution across habitat
covariates, we coarsened or interpolated all layers to a resolution of 250 m x
250 m. A detailed description of how we prepared each habitat covariate is
provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in ISSF models \citep{Avgar.2016, Fieberg.2021}. Movement
metrics were calculated for each step and included the step length
(\textsf{sl}), its natural logarithm (\textsf{log(sl)}), and the cosine of the
relative turning angle (\textsf{cos(ta)}). Moreover, we created the binary
variable \textsf{LowActivity} indicating whether a step was realized during
periods of low wild dog activity (09:00 to 17:00 local time) or high wild dog
activity (17:00 to 09:00 local time, \citealp{Cozzi.2012}). We performed all
data preparations, spatial computations, and statistical analysis in R, version
3.6.6 \citep{R.2020}. Some helper functions were written in {\tt C++} and
imported into {\tt R} using the {\tt Rcpp} package \citep{Eddelbuettel.2011,
Eddelbuettel.2013}.
